---
title: "Factors Affecting Students' Academic Performance"
author:
- 'Name: Ghulam Fatima'
- 'Roll No: BS-STATS-F25-R-36'
- 'Class: BS Statistics'
- 'Submitted to: Mam Wajiha Nasir'
date: '2026-06-11'
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## <u>Question 1:</u> Import Dataset and Display First Six Observations

```{r, comment = ""}
student_data <- read.csv("student_performance.csv")

knitr::kable(head(student_data))
```

**Interpretation:** The dataset is successfully loaded, and the head() function displays the first 6 rows to verify that all variable columns are correctly structured.

---

## <u>Question 2:</u> Determine Observations, Variables, and Names

```{r, comment = ""}
dim(student_data)

names(student_data)
```

**Interpretation:** The dataset contains exactly 50 student observations (rows) and 7 unique variables (columns) representing their academic metrics.

---

## <u>Question 3:</u>Descriptive Statistics of Quantitative Variables

```{r, comment = ""}
quantitative_data <- student_data[, 2:7]

knitr::kable(summary(quantitative_data))
knitr::kable(sapply(quantitative_data, sd), col.names = "Standard Deviation")
```

**Interpretation:** Students study an average of 22.2 hours per week and achieve an average final score of 74.2 marks. Attendance is generally high, indicating good class participation among students.

---

## <u>Question 4:</u>Identify Students with the Highest and Lowest Final Scores

```{r, comment = ""}
highest_score <- student_data[which.max(student_data$Final_Score), ]

lowest_score <- student_data[which.min(student_data$Final_Score), ]

extreme_students <- rbind(highest_score, lowest_score)
rownames(extreme_students) <- c("Highest Score", "Lowest Score")

knitr::kable(extreme_students)
```

**Interpretation:** The highest-scoring student achieved 95 marks, while the lowest-scoring student obtained 52 marks.

---


## <u>Question 5:</u> Identify Students with the Highest and Lowest Final Scores

```{r, comment = ""}
avg_study  <- round(mean(student_data[, 2]), 2)
avg_attend <- round(mean(student_data[, 3]), 2)
avg_score  <- round(mean(student_data[, 7]), 2)

average_table <- data.frame(
  Variable = c("Study Hours", "Attendance", "Final Score"),
  Average = c(avg_study, avg_attend, avg_score)
)

knitr::kable(average_table)
```

**Interpretation:** Students study an average of 22.2 hours per week, maintain 87.38% attendance, and achieve an average final score of 74.2 marks.

---

## <u>Question 6:</u> Histogram for Final Score

```{r, comment = ""}
hist(student_data[, 7], 
     main = "Histogram of Final Scores", 
     xlab = "Final Score (%)", 
     ylab = "Number of Students", 
     col = "lightblue", 
     border = "black")
```

**Interpretation:** The histogram indicates that most students obtained scores around the average range, with fewer students having very low or very high scores.

---

## <u>Question 7:</u> Boxplot Analysis of Study Hours

```{r, comment = ""}
boxplot(student_data[, 2], 
        main = "Boxplot of Study Hours", 
        ylab = "Weekly Study Hours", 
        col = "lightgreen", 
        border = "darkgreen")
```

**Interpretation:** The boxplot shows a clean distribution with no isolated dots, indicating that there are no outliers in Study Hours.

---

## <u>Question 8:</u>  Comparison of Final Scores for the First 10 Students

```{r, comment = ""}
first_10_students <- student_data[1:10, ]

barplot(first_10_students[, 7], 
        names.arg = first_10_students[, 1], 
        main = "Final Scores of First 10 Students", 
        xlab = "Student ID", 
        ylab = "Final Score (%)", 
        col = "lightcoral", 
        border = "black")
```

**Interpretation:** The bar chart compares the final scores of the first 10 students, showing individual variations in their academic performance.

---

## <u>Question 9:</u> Scatter Plot of Study Hours vs Final Score

```{r, comment = ""}
plot(student_data[, 2], student_data[, 7],
     main = "Scatter Plot of Study Hours vs Final Score",
     xlab = "Weekly Study Hours",
     ylab = "Final Score (%)",
     col = "blue",
     pch = 19)

abline(lm(student_data[, 7] ~ student_data[, 2]), col = "red", lwd = 2)
```

**Interpretation:** The scatter plot shows a strong positive relationship, meaning that as weekly study hours increase, the final scores of the students also increase.

---

## <u>Question 10:</u> Correlation Heatmap for Numerical Variables

```{r, fig.height=6, fig.width=6, comment = ""}
num_cols <- student_data[, 2:7]

corr_matrix <- cor(num_cols)

heatmap(corr_matrix, 
        Rowv = NA, Colv = NA, 
        main = "Correlation Heatmap", 
        col = heat.colors(100),
        cexCol = 0.8,
        cexRow = 0.8,
        margins = c(10, 10))
```

**Interpretation:** :  The heatmap indicates positive relationships among several academic variables. Study Hours, Attendance, and Previous GPA show strong positive associations with Final Score.

---

## <u>Question 11:</u> Simple Linear Regression Model

```{r, comment = ""}
simple_model <- lm(student_data[, 7] ~ student_data[, 2])

intercept <- round(coef(simple_model)[1], 2)
slope     <- round(coef(simple_model)[2], 2)

r_squared <- round(summary(simple_model)$r.squared, 4)

regression_results <- data.frame(
  Metric = c("Intercept (β₀)", "Slope (β₁)", "R2"),
  Value = c(intercept, slope, r_squared)
)

knitr::kable(regression_results)
```
**Predicted Final Score =** 43.92 + 1.36(Study Hours)
**Interpretation:** •  For every additional study hour, the final score increases by approximately 1.36 marks. 
•  The model explains approximately 99.26% of the variation in final scores.


---

## <u>Question 12:<u/> Multiple Linear Regression Model

```{r, comment = ""}
multiple_model <- lm(student_data[, 7] ~ student_data[, 2] + student_data[, 3] + student_data[, 6])

b0 <- round(coef(multiple_model)[1], 2)
b1 <- round(coef(multiple_model)[2], 2)
b2 <- round(coef(multiple_model)[3], 2)
b3 <- round(coef(multiple_model)[4], 2)

r_sq_multi <- round(summary(multiple_model)$r.squared, 4)

multi_results <- data.frame(
  Coefficient = c("Intercept (β0)", "Study Hours (β₁)", "Attendance (β2)", "Previous GPA (β3)", "R2"),
  Value = c(b0, b1, b2, b3, r_sq_multi)
)

knitr::kable(multi_results)
```

**Predicted Final Score =** 27.37 + 0.77(Study Hours) + 0.08(Attendance) + 7.37(Previous GPA)
**Interpretation:** •  An increase of one study hour increases the final score by approximately 0.77 marks. 
•  A 1% increase in attendance increases the final score by approximately 0.08 marks. 
•  A one-point increase in GPA increases the final score by approximately 7.37 marks. 
•  The model explains approximately 99.55% of the variation in final scores.
